---
title: "Introduction to artbin"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Introduction to artbin}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
library(artbin)
```

`artbin` calculates the sample size or power for randomised trials with binary
outcomes.  It supports superiority, non-inferiority (NI), and
substantial-superiority designs; two-arm and multi-arm (k-group) trials; and a
range of options including continuity correction, unequal allocation, and
loss-to-follow-up adjustment.

The package is a faithful R translation of the Stata `artbin` package (v2.1.2)
described in:

> Marley-Zagar E, White IR, Royston P, Barthel F, Parmar MKB, Babiker AG.
> artbin: Extended sample size for randomised trials with binary outcomes.
> *Stata J* 2023; 23(1): 24–52.

The examples in this vignette reproduce the five examples from Section 3 of
that paper.

---

## 1  Superiority trial — Wald test (Pocock 1983)

Pocock (1983) calculated the sample size for a 2-arm superiority trial
comparing anturan with placebo in patients after myocardial infarction.  The
primary outcome was death from any cause within one year.  Control-arm event
probability $\pi_1 = 0.10$; treatment-arm $\pi_2 = 0.05$; two-sided
$\alpha = 0.05$; power $= 90\%$.

Because $\pi_2 < \pi_1$, the trial aims to *reduce* the event probability
(death), so the outcome is **unfavourable** — `artbin` infers this
automatically.  Pocock used the Wald test (sample-estimate null variance):

```{r pocock}
artbin(pr = c(0.1, 0.05), alpha = 0.05, power = 0.9, wald = TRUE)
```

The total sample size of **1156** (578 per arm) matches Pocock's published
value.

The same result is obtained by framing the outcome as survival
($\pi_1 = 0.90$, $\pi_2 = 0.95$) with `favourable = TRUE`:

```{r pocock-fav}
artbin(pr = c(0.9, 0.95), alpha = 0.05, power = 0.9, wald = TRUE,
       favourable = TRUE)
```

---

## 2  Score test and comparison with R's `prop.test`

The default test in `artbin` is the score test (Farrington-Manning constrained
maximum likelihood under $H_0$), the same test used by R's `pwr` package and
Stata's `power twoproportions`:

```{r score}
artbin(pr = c(0.1, 0.05), alpha = 0.05, power = 0.9)
```

The score test requires **1164** participants — slightly more than the Wald
test because the constrained-ML null variance is larger than the sample
estimate.

For reference, R's built-in `power.prop.test` gives the same result:

```{r prop-test}
power.prop.test(p1 = 0.1, p2 = 0.05, sig.level = 0.05, power = 0.9)
```

---

## 3  One-sided non-inferiority trial

A non-inferiority trial assumes that the new treatment is *at least as good* as
the control, within a specified margin.  Here both arms have an anticipated
survival probability of 90%, and the NI margin is 5 percentage points
($m = -0.05$ means treatment must be no more than 5 pp worse than control):

```{r ni}
artbin(pr = c(0.9, 0.9), margin = -0.05, onesided = TRUE)
```

The one-sided $\alpha = 0.05$ test requires **457 patients per arm** (914
total).

### Calculating power for a fixed sample size

Supply `n` instead of `power` to calculate power:

```{r ni-power}
artbin(pr = c(0.9, 0.9), margin = -0.05, onesided = TRUE, n = 914)
```

---

## 4  Multi-arm superiority trial

When three or more event probabilities are supplied, `artbin` uses a global
chi-squared test to assess whether any difference exists among groups.  The
sample size for a 4-arm trial ($\pi = 0.1, 0.2, 0.3, 0.4$) at
$\alpha = 0.10$, power $= 90\%$:

```{r kgroup}
artbin(pr = c(0.1, 0.2, 0.3, 0.4), alpha = 0.1, power = 0.9)
```

**44 patients per arm** (176 total) are required.

The return list includes per-arm sample sizes `n1`–`n4` and per-arm expected
events `D1`–`D4`:

```{r kgroup-arms}
r <- artbin(pr = c(0.1, 0.2, 0.3, 0.4), alpha = 0.1, power = 0.9)
data.frame(
  arm   = paste0("Arm ", 1:4),
  pr    = c(0.1, 0.2, 0.3, 0.4),
  n     = c(r$n1, r$n2, r$n3, r$n4),
  D     = c(r$D1, r$D2, r$D3, r$D4)
)
```

---

## 5  Complex NI trial — STREAM

The STREAM trial evaluated a shorter intensive treatment for multi-drug
resistant tuberculosis against the WHO-recommended lengthier regimen.  Key
design features:

- Favourable binary outcome (culture-negative at 132 weeks): $\pi_1 = 0.70$
  (control), $\pi_2 = 0.75$ (treatment)
- NI margin $m = -0.10$ (treatment must be no more than 10 pp worse)
- 1:2 allocation (twice as many in treatment)
- 20% loss to follow-up
- Wald test (standard for NI trials)

```{r stream}
artbin(pr = c(0.7, 0.75), margin = -0.1, wald = TRUE,
       aratios = c(1, 2), ltfu = 0.2)
```

**398 patients** are required (133 control, 265 treatment), after inflating for
20% LTFU.  The STREAM trial concluded that the shorter regimen was non-inferior
to the WHO standard.

---

## Additional options

### Continuity correction

Apply Fleiss's continuity correction with `ccorrect = TRUE`:

```{r ccorrect}
artbin(pr = c(0.1, 0.05), alpha = 0.05, power = 0.9, ccorrect = TRUE)
```

### Blackwelder NI reference

A standard Blackwelder (1982) NI calculation with Wald test, one-sided
$\alpha = 5\%$:

```{r blackwelder}
artbin(pr = c(0.1, 0.1), margin = 0.2, alpha = 0.1, power = 0.9, wald = TRUE)
```

### Trend test for ordered dose–response

For k-arm trials with a dose–response structure, `trend = TRUE` uses a linear
trend test instead of the global chi-squared:

```{r trend}
artbin(pr = c(0.2, 0.3, 0.4), trend = TRUE)
```

### Unrounded sample sizes

`noround = TRUE` returns exact (unrounded) per-arm sizes, useful for
sensitivity analysis or when you want to apply your own rounding:

```{r noround}
artbin(pr = c(0.1, 0.05), aratios = c(1, 2), noround = TRUE)
```

---

## References

Blackwelder, W.C. (1982). "Proving the null hypothesis" in clinical trials.
*Controlled Clinical Trials*, 3(4), 345–353.

Farrington, C.P. and Manning, G. (1990). Test statistics and sample size
formulae for comparative binomial trials with null hypothesis of non-zero risk
difference or non-unity relative risk. *Statistics in Medicine*, 9(12),
1447–1454.

Nunn, A.J., et al. (2019). A trial of a shorter regimen for rifampin-resistant
tuberculosis. *New England Journal of Medicine*, 380(13), 1201–1213.

Pocock, S.J. (1983). *Clinical Trials: A Practical Approach*. Wiley.
